Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-18 talk/pepm2011/paper.tex
small things, more fundamental changes to come

\subsection{Static Objects}

The main insight to improve the code shown in the last section is that objects
in category 1 don't survive very long -- they are used only inside the loop and
nobody else in the program stores a reference to them. The idea for improving
the code is to analyze which objects fall in category 1 and therefore do
not have to be allocated at all.

This is a process that is usually called \emph{escape analysis}. In this paper we will
perform escape analysis by using partial evaluation. The partial evalution is a
bit peculiar in that it receives no static input arguments for the trace,
but it is only used to optimize operations within a trace.

The partial evaluation works by traversing the trace from beginning to end.
Whenever a \lstinline{new} operation is seen, the operation is removed and a \emph{static
object}\footnote{Here ``static'' is meant in the sense of partial
evaluation, \ie known at partial evaluation time, not in the sense of static
allocation or static method.} is constructed and associated with the variable
that would have stored
the result of \lstinline{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \lstinline{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\lstinline{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\lstinline{guard_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type and thus the outcome of
the type check the guard does is statically known.

>>>>

The main insight to improve the code shown in the last section is that objects
in category 1 do not survive very long -- they are used only inside the loop and
there is no other outside reference to them. The idea for improving
the code is to analyze which objects fall in category 1 and therefore do
not have to be allocated at all.

This is a process that is usually called \emph{escape analysis}
\cite{goldberg_higher_1990}. In this paper we will
perform escape analysis by using partial evaluation. The use of partial evaluation is a
bit peculiar in that it receives no static input arguments for the trace,
but it is only used to optimize operations within the trace.

To optimize the trace, it is traversed from beginning to end.
Whenever a \lstinline{new} operation is seen, the operation is removed and a \emph{static
object}\footnote{Here ``static'' is meant in the sense of partial
evaluation, \ie known at partial evaluation time, not in the sense of static
allocation or static method.} is constructed and associated with the variable
that would have stored
the result of \lstinline{new}. The static object describes the shape of the
original object, \eg where the values that would be stored in the fields of the
allocated object come from, as well as the type of the object. Whenever the
optimizer sees a \lstinline{set} that writes into such an object, that shape
description is updated and the operation can be removed, which means that the
operation was done at partial evaluation time. When the optimizer encounters a
\lstinline{get} from such an object, the result is read from the shape
description, and the operation is also removed. Equivalently, a
\lstinline{guard_class} on a variable that has a shape description can be removed
as well, because the shape description stores the type and thus the outcome of
the type check the guard does is statically known.
